Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-13 paper.tex
update intro

[TL:DR] Structure the introduction around this great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.

\subsection{ARGs are awesome, or general intro about approximations}

Knowledge of the exact genealogy of a set of samples would make certain problems 
in population genetics (e.g.\ inferring recombination rates) trivial. 
Multitude of methods have started to 
demonstrate the value of attempts to capture the details of these intricately 
interwoven paths of inheritance referred to as an ancestral recombination graph (ARG).
Recent advancements: the inference of ARGs at population scale.
Yet even though potential of ARGs in population and statistical genetics has been 
demonstrated it remains unclear how to connect these reconstructed ARGs to 
the stochastic/generative models that have lead us to study these datastructures 
in the first place.

[Message needs to be here: way to think about genealogies is tied to the CwR. However,
CwR does not enable us to simply compute a likelihood in the absence of a fully 
detailed history. The way forward instead has always been driven by approximations.]

% The coalescent with recombination 
% Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
% the generative model that allows us to understand the genealogy 
% underlying a particular sampleset. 
% Datastructure that captures it: ARG.

@@

% There is a discrepancy between ARGs as object of theoretical reasoning 
% vs what can be known. Explain the difference between precision/detail and accuracy. 
% High precision can give a false sense of certainty in the provided estimate.

% In population genetics, the questions we try to answer are generally 
% about the unobserved processes that gave rise to the genetic variation 
% we can observe given a set of samples.

\subsection{approximations have driven the rate of progress in inference}

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. Two approximations 
in particular have proven to be instrumental: the sequentially Markovian coalescent, 
and its extensions (SMC) on the one hand, and the Li and Stephens model on the other. 
The impact of both models is tightly linked to the fact that they have enabled 
the reformulation of inference problems as hidden Markov models (HMM) and thus 
the use of the well-known efficient HMM-algorithms.
%Both: 1/ Left to right, and 2/ discetization 
%of state space by emumerating tree topologies and/or discretising time.\\

@@

The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant  
leap in scalability. These methods derive their scalability at least to a certain 
degree from the fact that the ARGs they infer are approximate structures. In particular, 
recombination is an emergent property of the inferred ARGs explaining tree topology 
changes going from one local tree to the next. The recombination events themselves 
are not explicitly inferred and are thus not part of the resulting data structure.

\subsection{Recent progress in ARGs through approximate structures}

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].

Downsides:
Point estimate:
Some approaches have embraced the approximate nature of the resulting ARG by, for 
example, highlighting the inability to resolve certain splits and instead representing 
them as polytomies \citep{kelleher_2019_inferring}. 
+
This means that the degree of 
completeness in which genetic inheritance from ancestors to descendants is documented 
by each of these methods varies extensively.

But: Advantage of being able to infer ARGs at scale.

Connect this to likelihood approach.

While this is not a problem when describing the results of simulations 
(where all details are perfectly known), it is a major issue when we wish to formally 
describe the output of inference methods, particularly as datasets approach the population 
scale. 
Only successful if we can connect these to statistical models.

Progress driven by approximation: here similarly

Note that approximate structure need not be an issue, the fully precise and detailed 
ARG can never be known. Instead, the main question is how to connect these 
approximate structures to existing models.

\subsection{The Big problem: how do you connect such approximate structures to a generative model}

This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference.
The problem is that there is no direct way of calculating the coalescent 
likelihood from a set of marginal genealogies (i.e. integrating over the 
possible ARGs that could have given the set of marginal genealogies).

\subsection{what we will do}

Here, the idea is to use the backwards-in-time formulation of the SMC to 
compute likelihoods for for \textit{any} inferred ARG 
by integrating out the time to recombination.
The application of the SMC in a way that resonates more 
with the true direction of flow of genetic information 
results in a concrete way to formulate an algorithm 
that is \textit{general} in the sense that it can deal with the various levels of 
completeness that characterise the current state of the field of ARG reconstruction.

@@

[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.\\

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \tskit library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strenghtens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
[NOTE: these options have already been implemented in the current implementation]

@@

[TO DO]: improve results sections with some good viz,
We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
After removing the recombination nodes the likelihoods as defined here were computed both with and 
without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
explained is higher when the unary nodes are retained.

@@

[ TO DO final conclusion]

>>>>

[inspiration] great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.

[TL;DR]: way to think about genealogies is tied to the CwR. However,
CwR does not enable us to simply compute a likelihood in the absence of a fully 
detailed history. In the past, this has given rise to approximations to the 
coalescent that had similar properties yet were significantly easier to use 
as a basis for inference. 
A similar process has taken place in ARG-based inference, 
where scalable ARG inference has been made possible due to both these approximations 
to the coalescent as well as the idea of ARGs without explicit, fully detailed 
recombination histories. What is left however is to connect the two.]

%% ARGs are awesome, or general intro about approximations

[FIX ME: start is a mess. Need to find the right way in.]
Knowledge of the exact genealogy of a set of samples would make certain problems 
in population genetics (e.g.\ inferring recombination rates) trivial. 
Multitude of methods have started to 
demonstrate the value of attempts to capture the details of these intricately 
interwoven paths of inheritance referred to as an ancestral recombination graph (ARG).
Recent advancements: the inference of ARGs at population scale.
Yet even though potential of ARGs in population and statistical genetics has been 
demonstrated it remains unclear how to connect these reconstructed ARGs 
(as being approximate structures) to 
the stochastic/generative models that have lead us to study these datastructures 
in the first place.

@@

%% approximations have driven the rate of progress in inference
Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. Two approximations 
in particular have proven to be instrumental: the sequentially Markovian coalescent, 
and its extensions (SMC) on the one hand, and the Li and Stephens model on the other. 
The impact of both models is tightly linked to the fact that they have enabled 
the reformulation of inference problems as hidden Markov models (HMM) and thus 
the use of the well-known efficient HMM-algorithms.

@@

%% Recent progress in ARGs through approximate structures}

The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant  
leap in scalability. These methods derive their scalability at least to a certain 
degree from the fact that the ARGs they infer are approximate structures. In particular, 
recombination is an emergent property of the inferred ARGs, explaining tree topology 
changes going from one local tree to the next. The recombination events themselves 
are not explicitly inferred and are thus not part of the resulting data structure.

The computational gains from not having to pinpoint recombination events tie in 
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS]. 
Some approaches have embraced the approximate nature of the resulting ARG by, for 
example, highlighting the inability to resolve certain splits and instead representing 
them as polytomies \citep{kelleher_2019_inferring}. Taken together this means that the 
degree of completeness in which genetic inheritance from ancestors to descendants is documented 
by each of these methods varies extensively.
[TO ADD: these methods only provide point estimates]

%% The Big problem: how do you connect such approximate structures to a generative model

This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based description that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference.

%% what we will do
Here, the idea is to bridge this gap and use the backwards-in-time formulation of the SMC 
to compute likelihoods for for \textit{any} inferred ARG 
by integrating out the time to recombination.
The application of the SMC in a way that resonates more 
with the true direction of flow of genetic information 
results in a concrete way to formulate an algorithm 
that is \textit{general} in the sense that it can deal with the various levels of 
completeness that characterise the current state of the field of ARG reconstruction.

@@

[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
Is there more to say them the simple remark I made in \ref{fig:smc-unary}?\\

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \tskit library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strenghtens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample (see \ref{fig:3-arg-slices}). 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
[NOTE: these options have already been implemented in the current implementation]

@@

[TO DO]: improve the (scatter) plots that are referred to in this section such that not 
all questions will be about why likelihood values do not exactly correspond. How to 
compare these likelihoods in a more meaningful way?
We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
After removing the recombination nodes the likelihoods as defined here were computed both with and 
without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
explained is higher when the unary nodes are retained.

@@

[ TO DO final conclusion?]
